\section{Related work}
\label{sec:related}

\paragraph{Program extraction.}  The erasure calculus, the treatment of
singleton eliminations, informative-singleton unboxing and
logical-argument pruning follow Letouzey
\cite{Letouzey2004,Letouzey2003,Letouzey2008}, whose extraction descends
from Paulin-Mohring's realizability-based extraction for the Calculus of
Constructions \cite{PaulinMohring1989}, itself in the tradition of
Kleene realizability \cite{Kleene1945,TroelstraVanDalen1988}.  Our
specification extraction $\guardf$ is precisely the structural
realizability predicate of \cite{PaulinMohring1989} and the simulation
predicate $\sem{\cdot}$ of \cite[\S 2.4]{Letouzey2004}, read as a
first-order formula about the extracted constant; the fundamental lemma
(\cref{lem:fundamental}) plays the role of extraction correctness.  The
differences are dictated by the target: extraction must protect
evaluation order (dummy abstractions, argument-absorbing fixpoints),
which an equational target does not care about, so erasure can prune
outright (\cref{sec:lambdabox}); conversely an equational target cannot
rely on lazy non-execution, which is why well-founded recursion is a
genuine boundary (\cref{sec:limitations-wf}) rather than a solved case.
Machine-checked accounts of erasure correctness exist in the MetaCoq
project \cite{SozeauEtAl2020}; formally connecting our $\erasef$ to
MetaCoq's verified erasure is an attractive future step.

\paragraph{Hammer translations.}  The immediate context is the
CoqHammer translation \cite{CzajkaKaliszyk2018,CzajkaKaliszyk2016} and
its proof-theoretically sound core, the shallow embedding of
proof-irrelevant PTSs into first-order logic \cite{Czajka2018}.
Relative to \cite{Czajka2018}, this paper adds what that paper
explicitly leaves out --- inductive types, pattern matching, recursion
and refinement types --- at the price of a model-theoretic rather than
proof-theoretic soundness statement, and of assuming a stratified
monomorphic fragment rather than an arbitrary logical piPTS.  Relative
to \cite{CzajkaKaliszyk2018}, the case-axiom and inhabitation-atom
layers are replaced by the extraction-factored design, whose possibility
is conjectured in \S 9 of that paper.  For HOL-based hammers the
corresponding translation stack --- $\lambda$-lifting, arity
optimisation, type guards and tags with monotonicity-based soundness
proofs --- is developed by Meng and Paulson \cite{MengPaulson2008} and
Blanchette et al.\ \cite{BlanchetteEtAl2016}; our guard predicates are
the unary (``featherweight'') end of that design space, and our
model-with-junk construction is the same proof pattern used there for
guard-omission soundness.  Sledgehammer's practice
\cite{PaulsonBlanchette2010,BlanchetteEtAl2016qed} supplies the
operational evidence for S5: HOL hammers export per-constructor
\texttt{simps} equations produced by the definitional packages
\cite{Krauss2010}, never disjunctive match axioms.  Early precedents for
encoding dependent products with type predicates in untyped first-order
logic go back to Tammet and Smith \cite{TammetSmith1998}; the
types-as-predicates discipline is also Mizar's \cite{Wiedijk2007}.

\paragraph{Dependent type theory to ATPs.}  Lean-auto
\cite{QianEtAl2025} translates Lean~4 goals through monomorphisation
into higher-order logic, deliberately consuming the elaborator's
\emph{equational lemmas} for defined functions instead of raw
match-trees --- independent confirmation that equations are the right
currency for provers, obtained there from the frontend, here from the
translation itself.  \v{S}inkarovs and Rawson's Agda--Vampire bridge
\cite{SinkarovsRawson2026} restricts to an equational-Horn fragment
shared by both systems, again matching the shape our compilation
produces.  The Equations package \cite{SozeauMangin2019} generates
defining equations and elimination principles inside Coq for its own
definitions; our compilation produces the analogous equations uniformly
for plain definitions, at the FOL level.

\paragraph{Semantics and termination.}  The proof-irrelevant,
classical interpretation of propositions is the standard
proof-irrelevant model construction for CIC-like theories
\cite{Werner1997,MiquelWerner2003}, cut down to a truth-value semantics
over a syntactic domain.  The rank-restricted interpretations used for
the fix case of \cref{lem:fundamental} are the semantic counterpart of
guarded recursion in the style of Giménez \cite{Gimenez1995} and of
type-based termination \cite{BartheEtAl2004}; inductive definitions by
stages are classical \cite{Aczel1977}.  The confluence proof of
\cref{sec:confluence} is Takahashi's method \cite{Takahashi1995};
orthogonality-based alternatives go back to Klop \cite{Klop1980} and
are textbook material \cite{BaaderNipkow1998}.
