\section{Shallowness}
\label{sec:shallow}

Shallowness is the design constraint that shaped every clause of
\cref{sec:translation}; this section makes it precise and verifies it.
The operational argument is the one articulated for the core embedding in
\cite[\S 1.1]{Czajka2018}: first-order ATPs are heavily optimised for the
primitives of first-order logic --- clausification, resolution,
superposition with term indexing, E-matching --- so a translation is
efficient to the extent that the logical structure of the source arrives
at the prover \emph{as} first-order structure.  A \emph{deep} embedding,
in which a source proposition $\tau$ becomes a first-order \emph{term}
$\compf(\tau)$ related to its proofs by an inhabitation predicate
$\mathsf{T}(x, \compf(\tau))$, forces every logical inference of the
object logic to be simulated by instantiations of bridge axioms, and
conjectures become existence statements ($\exists y.\,\mathsf{T}(y,
\compf(\tau'))$) that require the prover to synthesise a term standing
for a proof --- both dramatically hostile to first-order proof search
\cite[\S 1.1]{Czajka2018}.

Concretely, we verify the following properties of the translation of
\cref{sec:translation}.

\begin{description}[leftmargin=2em]
\item[S1 (Native logical structure).]  $\Ff$ maps every connective and
  quantifier of a shallow proposition to the corresponding first-order
  connective or (guarded) quantifier, and source equality to first-order
  equality $\feq$, \emph{at translation time} (\cref{def:guard-F}).  No
  logical structure of a shallow proposition is reified into a term.
  Consequence: the clausal form of $\F{\varphi}$ is the clausal form one
  would write by hand for $\varphi$.

\item[S2 (No inhabitation predicate).]  $\SigFOL$ contains no binary
  ``has-type'' predicate and no term encoding of types
  (\cref{def:signature}).  The only typing-flavoured atoms are the unary
  guards $\Ihat{I}(u)$, occurring exactly at atomic-type leaves of
  $\guardf$ --- the soft-typing base case, structurally the device used
  by Mizar's type predicates \cite{Wiedijk2007} and by the type guards
  of HOL hammers \cite{MengPaulson2008,BlanchetteEtAl2016}.

\item[S3 (Per-occurrence payload expansion).]  The
  propositional content of a refinement type is expanded inline at every
  occurrence, in both polarities --- as a premise under $\forall$
  (negative position) and as a conjunct under $\exists$ and in
  conclusions (positive position) ---
  licensed by the fact that \cref{lem:coincidence} is an equivalence,
  not an implication.  There is no $\mathsf{sig}$ symbol, no boxing
  constructor, no projection function and no unfolding axiom in the
  output: nothing is left for the ATP to unfold.  This is the clause
  that repairs the local relapse into deep embedding identified in
  \cref{sec:intro} --- compare the specification axiom of \cref{ex:h}
  with the opaque $\mathsf{HasType}(h, \mathsf{sig}(\dots))$ atom it
  replaces.

\item[S4 (Proofs never become terms).]  Erasure drops proof arguments
  and abstractions entirely (\cref{def:erasure}); the specification
  clause $\guard{\varphi \to \tau}{u} = \F{\varphi} \to \guard{\tau}{u}$
  keeps the first-order term $u$ unapplied across a propositional
  premise, in lock-step with the pruned arity of the extracted program.
  The prover consequently never constructs, matches or instantiates any
  term standing for a proof.  (The dead-code constant $\prfc$ can occur,
  but never as an argument at a pruned position --- only as the image of
  $\bot$-elimination.)

\item[S5 (Definitions as oriented computation).]  The definitional
  content of the environment arrives as per-constructor pattern
  equations (\cref{def:compile}) --- orientable unit or near-unit
  equations forming an orthogonal rewrite system
  (\cref{rem:eq-shape}) --- rather than as guarded disjunctions
  (\cref{sec:casesplit-cnf}).  Case analysis (a search operation) is
  segregated into the per-datatype inversion axioms, so provers pay for
  splitting only when the proof actually requires it.
\end{description}

Two further points delimit what shallowness can and cannot mean here.

\paragraph{The residue.}  Full first-order structure cannot always be
achieved: partial applications go through the application operator
$\fapp$ and pointer constants (\cref{def:compC}) --- the standard
arity-optimisation residue of every HOL-to-FOL translation
\cite{MengPaulson2008} --- and anonymous $\lambda$-abstractions, matches
and fixpoints inside formulas are lifted to fresh symbols with defining
axioms, which is unavoidable in a language without binders.  Both
mechanisms are local: they affect the translation of \emph{terms}, never
the logical skeleton of a proposition.  Propositions outside $\Sfrag$
(predicate quantification) have no shallow first-order image at all and
are excluded rather than encoded; the implemented tool falls back to
opaque atoms for them, a fallback whose soundness is a separate
(deep-embedding) question not treated here.

\paragraph{Shallowness versus soundness.}  The two requirements pull in
the same direction in this design.  The inline expansion of refinement
payloads (S3) is justified by the model equivalence of
\cref{lem:coincidence}; the guard-free equations (S5) are not merely
sound but strictly stronger than their guarded counterparts
(\cref{prop:interderivable}) while remaining valid in $\ML$
(\cref{prop:split-valid}).  Where a guard is genuinely load-bearing ---
exhaustiveness in inversion axioms (\cref{prop:disj-guard}) --- the
translation keeps it.  The soundness proof thus doubles as a shallowness
audit: every remaining non-native atom ($\Ihat{I}$ guards, lifted
symbols, $\fapp$) is one the model construction shows to be necessary
for the statement it protects, and every removable layer of indirection
has been removed at translation time, which is the design maxim shared
with the instantiation-level optimisations of the implemented tool: work
done at translation time is work the ATP does not repeat at search time,
once per relevant inference.
