% Compile with XeTeX:  just  (or: xelatex main; bibtex main; xelatex main; xelatex main)
\documentclass[11pt]{article}

\usepackage[a4paper,margin=2.7cm]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=blue!50!black,citecolor=green!35!black,urlcolor=blue!50!black]{hyperref}
\usepackage[capitalize,noabbrev]{cleveref}

\input{macros}

\numberwithin{equation}{section}
\emergencystretch=1.5em

\title{Sound Monomorphisation and Monotonicity-Based Guard Erasure\\
for Hammers for Dependent Type Theory}
\author{}
\date{AI draft of \today}

\begin{document}

\maketitle

\begin{abstract}
Hammers connect proof assistants to automated theorem provers (ATPs)
by translating goals and selected premises into first-order logic.
For dependent type theory the standard translation, implemented in
CoqHammer, is an untyped encoding in which typing information is
retained through \emph{guards}: atoms $\hastype(x,\tau)$ expressing
that the individual $x$ has type $\tau$.  Guards make the encoding
sound but verbose, and quantification over types makes the resulting
problems hard for ATPs.  Two optimisations promise substantial
improvements: \emph{heuristic monomorphisation}, which instantiates
type quantifiers with ground types occurring in the problem, and
\emph{guard erasure}, which deletes guards that provably do not
affect provability.  Both have long been used for simple type theory
in Sledgehammer-style tools, where their correctness rests on the
model theory of many-sorted first-order logic.  No comparable
analysis exists for dependent type theory; the standard soundness
proof for the CoqHammer translation is proof-theoretic and, as its
author observed, incompatible with the model-theoretic methods
underlying monotonicity inference.

This paper closes that gap.  We define a calculus $\CICm$ --- a
predicative fragment of the Calculus of Inductive Constructions with
an impredicative universe of propositions and parametric inductive
types --- together with a set-theoretic semantics, and give a
self-contained, model-theoretic soundness proof for a guarded
translation of $\CICm$ into classical untyped first-order logic: the
guard predicate is interpreted as set membership, and every
first-order consequence of the translated problem is true in the
induced standard model.  On this basis we prove three groups of
results.  First, heuristic monomorphisation --- formulated as
premise instantiation at the level of $\CICm$, followed by
normalisation and retranslation --- preserves soundness
unconditionally, and, when generic premises are kept, provability;
we also show that no complete recursive monomorphisation procedure
exists.  Second, \emph{signature specialisation}, the per-occurrence
replacement of guard, application and truth atoms by fresh
specialised symbols (without bridging axioms), preserves soundness,
and is provability-preserving on fully monomorphised problems.
Third, we formulate \emph{monotonicity-based guard erasure} for the
resulting monomorphic problems and prove it sound: guards of a type
$\tau$ may be erased whenever $\tau$ is witnessed inhabited and
either no variable of type $\tau$ occurs naked in a positive
equation, or the problem forces $\tau$ to be infinite --- a condition
that for inductive types is established by the injectivity and
discrimination axioms the translation itself emits.  We prove that
both side conditions are necessary, and that the classical
counterexample to guard erasure in the CoqHammer literature is
precisely a violation of monotonicity.  All proofs are given in
full.
\end{abstract}

\tableofcontents

\input{sec-intro}
\input{sec-calculus}
\input{sec-semantics}
\input{sec-translation}
\input{sec-mono}
\input{sec-specialization}
\input{sec-erasure}
\input{sec-pipeline}
\input{sec-related}
\input{sec-conclusion}

\bibliographystyle{alpha}
\bibliography{refs}

\end{document}
