\section{Tightness of the criteria}
\label{sec:tightness}

Each clause of the criteria excludes something, and each exclusion is
necessary: relaxing it admits a consistent environment whose transformed
translation refutes a goal not entailed by the source theory.  We record
one counterexample per exclusion.  In all four propositions ``unsound''
means: the transformed problem
$\Theta' \cup \Comp \cup \{\neg G'\}$ is unsatisfiable although
$\Th \not\entg \phi_0$ --- the negation of the guarantee provided by
\cref{thm:guards-sound,thm:erasure-soundness,thm:composition}.

\begin{proposition}[Equations do not cover: uncovered guards cannot be
  dropped]
  \label{prop:tight-uncovered}
  Let $\Env_1$ have inductive types $\mathit{unit}$ (constructor
  $\mathit{tt}$) and $\mathit{bool}$ (constructors
  $\mathit{true}, \mathit{false}$), no functions or predicates, and the
  single premise $L = \forall x{:}\mathit{unit}.\ x \eqx{\mathit{unit}}
  \mathit{tt}$.  Let $\phi_0 = \mathit{true} \eqx{\mathit{bool}}
  \mathit{false}$.  Then $\Env_1$ is consistent,
  $\ThOf{\Env_1} \not\entg \phi_0$, and replacing $\enc{L} =
  \forall x.\ \T(x, \widehat{\mathit{unit}}) \imp x = \hat{\mathit{tt}}$
  by the unguarded $\forall x.\ x = \hat{\mathit{tt}}$ makes the problem
  unsatisfiable.
\end{proposition}

\begin{proof}
  Consistency and non-entailment: the ground structure $\NN$ with
  $\NN_{\mathit{unit}} = \{\bullet\}$,
  $\NN_{\mathit{bool}} = \{0,1\}$ and the evident constructor
  interpretations satisfies $\ThOf{\Env_1}$ (injectivity, distinctness
  and exhaustiveness hold in the standard interpretations; $L$ holds) and
  falsifies $\phi_0$.  Unsatisfiability of the transformed problem: from
  $\forall x.\ x = \hat{\mathit{tt}}$ instantiate $x$ with
  $\hat{\mathit{true}}$ and $\hat{\mathit{false}}$ (both closed terms) to
  get $\hat{\mathit{true}} = \hat{\mathit{tt}} =
  \hat{\mathit{false}}$; the discrimination axiom
  $\hat{\mathit{true}} \neq \hat{\mathit{false}}$ of $\mathrm{ID}$
  yields $\bot$, so the problem is unsatisfiable regardless of the goal.
  The criterion of \cref{def:jtjf} rejects the step: the body
  $x = \hat{\mathit{tt}}$ is an equation, $\tc{x = \hat{\mathit{tt}}} =
  \emptyset$ (\cref{def:tc}), and no rule applies.  Hence the clause
  $\tc{\text{equation}} = \emptyset$ cannot be relaxed to let a variable's
  occurrence in an equation count as a cover.
\end{proof}

This is the fragment form of the classic finite-exhaustion unsoundness of
unguarded translations \cite{MengPaulson2008,CzajkaKaliszyk2018}: a
one-element type exports its extensionality to the whole untyped domain.

\begin{proposition}[Existential guards need $\jf$, not $\jt$]
  \label{prop:tight-exists}
  Let $\Env_2$ have the single inductive type $\mathit{void}$ with no
  constructors and no premises, and let
  $\phi_0 = \exists x{:}\mathit{void}.\ \top$.  Then $\Env_2$ is
  consistent, $\ThOf{\Env_2} \not\entg \phi_0$, and rewriting the goal
  translation $\enc{\phi_0} = \exists x.\ \T(x,\widehat{\mathit{void}})
  \wedge \top$ to $\exists x.\ \top$ --- which the \emph{universal}
  judgment would license, since $\jtx{x}{\widehat{\mathit{void}}}(\top)$
  holds --- makes the problem unsatisfiable.
\end{proposition}

\begin{proof}
  The ground structure with $\NN_{\mathit{void}} = \emptyset$ satisfies
  $\ThOf{\Env_2}$ --- exhaustiveness for $\mathit{void}$ is
  $\forall z{:}\mathit{void}.\ \bot$, vacuously true --- and falsifies
  $\phi_0$.  The transformed problem contains
  $\neg \exists x.\ \top$, which is unsatisfiable on its own (domains are
  nonempty).  The criterion rejects the step:
  \cref{thm:guard-equivalence}(ii) demands
  $\jfx{x}{\widehat{\mathit{void}}}(\top)$, and no rule derives
  $\jf(\top)$.  Hence the $\forall$/$\exists$ asymmetry of
  \cref{thm:guard-equivalence} is essential: junk-truth of the body makes
  a guarded \emph{universal} redundant but makes a guarded
  \emph{existential} strictly stronger than its unguarded form.
\end{proof}

\begin{proposition}[Result-only type arguments are not erasable]
  \label{prop:tight-nil}
  Let $\Env_3$ have inductive types $\mathit{nat}$ (constructors
  $\mathit{zero}$, $\mathit{succ}$), $\mathit{unit}$, $\mathit{bool}$
  (as above), a function constant
  $\mathit{card} : \forall\alpha.\ () \arr \mathit{nat}$ (no term
  arguments), and premises
  \[
    L_1 = \mathit{card}\langle\mathit{unit}\rangle()
          \eqx{\mathit{nat}} \mathit{succ}(\mathit{zero}),
    \qquad
    L_2 = \mathit{card}\langle\mathit{bool}\rangle()
          \eqx{\mathit{nat}} \mathit{zero} .
  \]
  Let $\phi_0 = \mathit{succ}(\mathit{zero}) \eqx{\mathit{nat}}
  \mathit{zero}$.  Then $\Env_3$ is consistent,
  $\ThOf{\Env_3} \not\entg \phi_0$, and erasing $\mathit{card}$'s type
  argument
  --- forbidden by \cref{def:erasable}, as position $1$ occurs in no term
  argument type --- makes the problem unsatisfiable.
\end{proposition}

\begin{proof}
  Consistency: the standard structure with
  $\mathit{card}_{\mathit{unit}} = 1$,
  $\mathit{card}_{\mathit{bool}} = 0$ (and standard $\mathit{nat}$)
  satisfies both premises and the structural axioms and falsifies
  $\phi_0$.  After the illegal erasure, both premises speak about the
  single constant $\hat{\mathit{card}}_\eps$:
  $\hat{\mathit{card}}_\eps = \hat{\mathit{succ}}(\hat{\mathit{zero}})$
  and $\hat{\mathit{card}}_\eps = \hat{\mathit{zero}}$, whence
  $\hat{\mathit{succ}}(\hat{\mathit{zero}}) = \hat{\mathit{zero}} =
  \enc{\phi_0}$; the problem containing $\neg\enc{\phi_0}$ is
  unsatisfiable.  (Where the soundness proof would break:
  \cref{lem:inference-unique} is inapplicable --- no term position
  witnesses $\alpha$ --- so \cref{def:MNe} has no well-defined proper
  clause for $\hat{\mathit{card}}_\eps$, and any total choice falsifies
  $L_1$ or $L_2$ in the erased model.)  This is the formal core of the
  $\mathit{card}\ \mathsf{UNIV}$ unsoundness of
  \cite{MengPaulson2008} and of the $\mathit{nil}$ example: erasing
  $\mathit{nil}$'s argument likewise conflates instances, with the
  collapse surfacing through premises such as
  $\mathit{rev}\langle\alpha\rangle(\mathit{nil}\langle\alpha\rangle)
  \eqx{} \mathit{nil}\langle\alpha\rangle$ at two instances.
\end{proof}

\begin{proposition}[Erasing predicate type arguments breaks covers]
  \label{prop:tight-pred-erasure}
  Let $\Env_4$ have inductive types $\mathit{nat}$, $\mathit{bool}$ (as
  above) and $\mathit{list}$ (constructors
  $\mathit{nil} : \forall\alpha.\ () \arr \mathit{list}(\alpha)$,
  $\mathit{cons} : \forall\alpha.\ \alpha \times \mathit{list}(\alpha)
  \arr \mathit{list}(\alpha)$), a predicate
  $q : \forall\alpha.\ \alpha \arr \PProp$, and premises
  \begin{align*}
    P_1 &= \forall l{:}\mathit{list}(\mathit{nat}).\
      q\langle \mathit{list}(\mathit{nat})\rangle(l) \imp
      l \eqx{\mathit{list}(\mathit{nat})} \mathit{nil}\langle
      \mathit{nat}\rangle(), \\
    P_2 &= q\langle \mathit{list}(\mathit{bool})\rangle\bigl(
      \mathit{cons}\langle\mathit{bool}\rangle(\mathit{true},
      \mathit{nil}\langle\mathit{bool}\rangle())\bigr).
  \end{align*}
  Let $\phi_0 = \mathit{true} \eqx{\mathit{bool}} \mathit{false}$.  Then
  $\Env_4$ is consistent, $\ThOf{\Env_4} \not\entg \phi_0$, and the
  combination
  of (a) erasing $q$'s type argument --- forbidden, since
  \cref{def:erasable} restricts $\Er$ to $\Constr \cup \Fun$ --- and
  (b) dropping the guard of $P_1$'s translation on the strength of the
  ``cover'' $\hat q_\eps(l)$ makes the problem unsatisfiable.
\end{proposition}

\begin{proof}
  Consistency: interpret $q_{\mathit{list}(\mathit{nat})}$ as
  $\{\mathit{nil}\}$, $q_{\mathit{list}(\mathit{bool})}$ as
  $\{[\mathit{true}]\}$, all other instances as empty, over the standard
  list structures; $P_1$, $P_2$ and the structural axioms hold, and
  $\phi_0$ fails.  After the two illegal steps the problem contains
  \[
    \forall l.\ \hat q_\eps(l) \imp l =
    \hat{\mathit{nil}}(\widehat{\mathit{nat}})
    \qquad\text{and}\qquad
    \hat q_\eps\bigl(\hat{\mathit{cons}}(\widehat{\mathit{bool}},
    \hat{\mathit{true}}, \hat{\mathit{nil}}(\widehat{\mathit{bool}}))\bigr).
  \]
  Instantiating the former with the latter's argument yields
  $\hat{\mathit{cons}}(\widehat{\mathit{bool}}, \hat{\mathit{true}},
  \hat{\mathit{nil}}(\widehat{\mathit{bool}})) =
  \hat{\mathit{nil}}(\widehat{\mathit{nat}})$, contradicting the
  unguarded discrimination axiom
  $\forall \alpha\, x\, l\, \alpha'.\
  \hat{\mathit{cons}}(\alpha,x,l) \neq \hat{\mathit{nil}}(\alpha')$ of
  $\mathrm{ID}$; the problem is unsatisfiable regardless of the goal.
  Where the soundness proof breaks: after erasing $q$, no completion
  axiom of the form
  $\forall \alpha\, x.\ \hat q_\eps(x) \imp \T(x, \alpha\text{-dependent
  type})$ is even expressible --- the natural interpretation of
  $\hat q_\eps(d)$ is ``$d$ satisfies \emph{some} instance of $q$'',
  which forces no particular type --- so $\hat q_\eps(l)$ is not strict
  and covers nothing.  With guards intact (only step (a)), or with $q$
  unerased (only step (b), which is then licensed and sound by
  \cref{thm:composition}), no contradiction arises.
\end{proof}

\begin{remark}
  The four propositions delimit the design space sharply: covers must be
  atoms whose predicate is \emph{false-extended} in the intended model
  (equations and non-rigid $\T$-atoms are not), existential guards are
  facts rather than obligations, type-argument recovery needs a witness
  among the \emph{retained term arguments}, and the two optimizations
  compose only because they partition the signature (predicates feed
  guard omission; functions and constructors feed erasure).  Every known
  practical unsoundness pattern of unguarded or type-erased hammer
  translations that falls inside our fragment is an instance of one of
  the four \cite{MengPaulson2008,BlanchetteBohmePopescuSmallbone2016,
  CzajkaKaliszyk2018}.
\end{remark}
